On the presentation of orthogonal groups over polynomial rings

### 2019-05-13 dl-paper.tex
Improved typesetting of Andrei's piece

For a simply-laced root system $\Phi$ and a commutative ring $R$, the Steinberg group $\St(\Phi,\,R)$ is generated by the formal symbols $x_{\alpha}(a)$ for $\alpha\in\Phi$, $a\in R$ subject to the Steinberg relations
\setcounter{equation}{0}
\renewcommand{\theequation}{R\arabic{equation}}
\begin{align}
&x_{\alpha}(a)x_{\alpha}(b)=x_{\alpha}(a+b),\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=x_{\alpha+\beta}(N_{\alpha\beta}\,ab),\text{ for }\alpha+\beta\in\Phi,\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\text{ for }\alpha+\beta\not\in\Phi\cup0.
\end{align}

Let $I\trianglelefteq A$ be an ideal in a commutative ring $A$ (not neccesarily proper), denote $R=\oplus_{d\in\mathbb Z}R_d\subseteq A[t,t\inv]$, where $R_d=It^d$, $d<0$, and $R_d=At^d$, $d\geq0$.

@@

Define trancated Steinberg groups $G_n$ for $n\geq1$ by the set of generators $x_\alpha(a)$, $\alpha\in\rD_l$, $a\in R_d$, $d\leq n$, and relations

\vspace{-15pt}
\setcounter{equation}{0}
\renewcommand{\theequation}{R\arabic{equation}$\!\,_n$}
\begin{align}
&\,\,x_{\alpha}(a)x_{\alpha}(b)=x_{\alpha}(a+b),\ a,b\in R_d,\ d\leq n,\\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]=x_{\alpha+\beta}(N_{\alpha\beta}\,ab),\ a\in R_d,\ b\in R_e,\ d,e,d+e\leq n,\phantom{tttttttt}
\end{align}
\vspace{-35pt}
\setcounter{equation}{0}
\renewcommand{\theequation}{R3.\arabic{equation}$\!\,_n$}
\begin{align}
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\ \alpha+\beta\not\in\rC_l\cup0,\ a\in R_d,\ b\in R_e,\ d,e\leq n,\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\ \alpha+\beta\in\rC_l\setminus\rD_l,\ a\in R_d,\ b\in R_e,\ d,e,d+e\leq n.
\end{align}

\begin{rem}
We can take $\mathbb R^l$ with a base $\chi_1,\ldots,\chi_l$ and visualise root systems
$$
\rD_l=\{\pm\chi_i\pm\chi_j\mid1\leq i\neq j\leq l\}
$$
and $\rC_l=\rD_l\cup\{\pm2\chi_i\mid1\leq i\leq l\}$. Then we immediately see that $\alpha+\beta\in\rC_l\setminus\rD_l$ exactly means that either $\{\alpha,\beta\}$ or $\{-\alpha,-\beta\}$ equals $\{\chi_i-\chi_j,\chi_i+\chi_j\}$. Then for some $\sigma$ from the Weyl group of $\rD_l$ we get 
$$
\{\sigma\alpha,\sigma\beta\}=\{\chi_{l-1}-\chi_l, \chi_{l-1}+\chi_l\}=\{\alpha_{l-1},\alpha_l\}
$$ 
(the horns of the Dynkin diagram $\rD_l$). Another condition on $\alpha$ and $\beta$ equivalent to $\alpha+\beta\in\rC_l\setminus\rD_l$ is that they do not lie in any subsystem of $\rD_l$ of type $\rA_m$ for $m\geq4$.
\end{rem}

@@

\begin{lemma}
\label{welldef}
Take $d$, $e\leq n$, $d+e=n+1$, $a\in R_d$, $b\in R_e$, and consider an embedding $\iota\colon\rA_4\hookrightarrow\rD_l$. For the chosen simple roots $\alpha_k$ of $\rA_4$ denote $\beta_k=\iota(\alpha_k)$. Then holds
$$
[x_{\beta_1}(a),\,x_{\beta_2+\beta_3}(b)]=[x_{\beta_1+\beta_2}(Nabt\inv),\,x_{\beta_3}(t)],
$$
where $N=N_{\beta_1,\beta_2}N_{\beta_2,\beta_3}$.
\end{lemma}
\begin{proof}
Observe that $d$, $e\geq1$. Take $x=x_{\beta_1}(a)$, $y=x_{\beta_2}(N_{\beta_3,\beta_2}bt\inv)$, $z=x_{\beta_3}(t)$. Since $x$ and $z$ lie in the subsystem of type $\rA_4$, we get by R3.1$\!\,_n$ that they commute. Then Hall--Witt identity gives us
$$
\,^y[x,\,[y\inv,\,z]]=\,^x[[x\inv,\,y],\,z],
$$
where $[x,\,[y\inv,z]]=[x_{\beta_1}(a),\,x_{\beta_2+\beta_3}(b)]$ and $[[x\inv,y],\,z]=[x_{\beta_1+\beta_2}(Nabt\inv),\,x_{\beta_3}(t)]$ by R2$\!\,_n$. Here $x$ commutes with $x_{\beta_1+\beta_2}(Nabt\inv)$ and $x_{\beta_3}(t)$, and $y$ commutes with $x_{\beta_2+\beta_3}(b)$ by R3.1$\!\,_n$. Finally, $[x_{\beta_1}(-a),\,y]=x_{\beta_1+\beta_2}(Nabt\inv)$ also commutes with $x_{\beta_2+\beta_3}(b)$, what finishes the proof.
\end{proof}

@@

Moreover, with the assumption of Lemma~\ref{welldef} on $a$ and $b$ we have
$$
[x_{\beta}(a),\,x_{\gamma}(b)]=s_{\alpha}(N_{\beta,\gamma}\,ab).
$$

The following observation is obvious.

\begin{lemma}
\label{commute}
Consider an embedding $\iota\colon\rA_4\hookrightarrow\rD_l$ and denote $\beta_k=\iota(\alpha_k)$ for simple roots $\alpha_k$ of $\rA_4$. Then for any $a\in R_d$, $b\in R_e$, $c\in R_f$ with $d$, $e$, $f\leq n$ holds
$$
[[x_{\beta_1}(a),\,x_{\beta_2}(b)],\,x_{\beta_4}(c)]=1.
$$
\end{lemma}

Consider now any two roots $\alpha$, $\beta\in\rD_l$ such that $\alpha+\beta\not\in\rC_l\cup0$. Since $l\geq5$, there exists a subsystem $\Psi$ of type $\rA_4$, containing $\alpha$ and $\beta$. Then we can take any $a\in R_{n+1}$ and $b\in R_d$ for $d\leq n$ and apply Lemma~\ref{commute} for $\Psi$ to get
\setcounter{equation}{0}
\renewcommand{\theequation}{\arabic{equation}}
\begin{align}
\label{sx}
[s_\alpha(a),\,x_\beta(b)]=1.
\end{align}
Next, using this fact we can apply again Lemma~\ref{commute} for $\Psi$ and get
$$
[s_\alpha(a),\,s_\beta(c)]=1
$$
for any $c\in R_{n+1}$. Therefore, R3.1$\!\,_{n+1}$ holds for $s_\alpha$'s.

Now one can take $a\in R_d$ for $d\leq0$ and $b\in R_{n+1}$, and repeat the proof of Lemma~\ref{welldef} using~(\ref{sx}) to get
$$
[x_\beta(a),\,s_\gamma(b)]=\begin{cases}s_{\beta+\gamma}(N_{\beta\gamma}\,ab),&d=0,\\x_{\beta+\gamma}(N_{\beta\gamma}\,ab),&d<0.\end{cases}
$$
This finishes the proof of R2$\!\,_{n+1}$.

@@

\begin{lemma}
For $a\in R_d$, $b\in R_e$, $d$, $e\leq n$, $d+e=n+1$ holds
$$
[x_{\alpha_{l-1}}(a),\,x_{\alpha_l}(b)]=1.
$$
\end{lemma}
\begin{proof}
Take $x=x_{\alpha_{l}-\alpha_{l-2}}(N_{\alpha_{l}-\alpha_{l-2},\alpha_{l-2}}bt\inv)$, $y=x_{\alpha_{l-1}-\alpha_{l-2}}(N_{\alpha_{l-2},\alpha_{l-1}-\alpha_{l-2}}at\inv)$, and $z=x_{\alpha_{l-2}}(t)$. Then
$$
[x_{\alpha_{l-1}}(a),\,x_{\alpha_l}(b)]=[[y\inv,\,z],\,[x,\,z]]=\,^{y\inv z}[y,\,[z\inv,\,x]]\cdot[y\inv,\,[x,\,z]].
$$
Next, by R3.2$\!\,_n$ we know that $x$ and $y$ commute, and therefore 
$$
\,^{y\inv z}[y,\,[z\inv,\,x]]=[[y\inv,\,z],\,x]=x_{\alpha_{l-1}+\alpha_{l}-\alpha_{l-2}}(Nabt\inv)
%[x_{\alpha_{l-1}}(a),\,x_{\alpha_{l}-\alpha_{l-2}}(N_{\alpha_{l}-\alpha_{l-2},\alpha_{l-2}}bt\inv)].
$$
for $N=N_{\alpha_{l-1},\alpha_{l}-\alpha_{l-2}}N_{\alpha_{l}-\alpha_{l-2},\alpha_{l-2}}$. Finally,
$$
[y\inv,\,[x,\,z]]=x_{\alpha_{l-1}-\alpha_{l-2}+\alpha_l}(N'abt\inv)
$$
for $N'=-N_{\alpha_{l-2},\alpha_{l-1}-\alpha_{l-2}}N_{{\alpha_{l-1}-\alpha_{l-2}},\alpha_l}$. Since $N_{\alpha,\beta}$ are the structure constants of the Lie algebra, it follows $N=-N'$ from the Jacobi identity.
\end{proof}

>>>>

For a simply-laced root system $\Phi$ and a commutative ring $R$, the Steinberg group $\St(\Phi,\,R)$ is generated by the formal symbols $x_{\alpha}(a)$ for $\alpha\in\Phi$, $a\in R$ subject to the Steinberg relations
\begin{align}
&x_{\alpha}(a)x_{\alpha}(b)=x_{\alpha}(a+b), \tag{R1}\\
&[x_{\alpha}(a),\,x_{\beta}(b)]=x_{\alpha+\beta}(N_{\alpha\beta}\,ab),\text{ for }\alpha+\beta\in\Phi, \tag{R2} \\
&[x_{\alpha}(a),\,x_{\beta}(b)]=1,\text{ for }\alpha+\beta\not\in\Phi\cup0. \tag{R3}
\end{align}

Let $I\trianglelefteq A$ be a (not neccesarily proper) ideal of a commutative ring $A$. Denote $R=\oplus_{d\in\mathbb Z}R_d\subseteq A[t,t\inv]$, where $R_d=It^d$, $d<0$, and $R_d=At^d$, $d\geq0$.

@@

Define truncated Steinberg groups $G_n$ for $n\geq1$ by the set of generators $x_\alpha(a)$, $\alpha\in\rD_l$, $a\in R_d$, $d\leq n$, and relations

\begin{align}
&\,\,x_{\alpha}(a)x_{\alpha}(b) =  x_{\alpha}(a+b),& a,b\in R_d,\ d\leq n, \tag{R$1_n$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= x_{\alpha+\beta}(N_{\alpha\beta}\,ab),& a\in R_d,\ b\in R_e,\ d,e,d+e\leq n, \tag{R$2_n$} \\
&[x_{\alpha}(a),\,x_{\beta}(b)]= 1,& \alpha+\beta\not\in\rC_l\cup0,\ a\in R_d,\ b\in R_e,\ d,e\leq n, \tag{R$3.1_n$} \\
&[x_{\alpha}(a),\,x_{\beta}(b)]= 1,& \alpha+\beta\in\rC_l\setminus\rD_l,\ a\in R_d,\ b\in R_e,\ d,e,d+e\leq n. \tag{R$3.2_n$}
\end{align}

\begin{rem}
We can take $\mathbb R^l$ with a base $\chi_1,\ldots,\chi_l$ and visualise root systems
\[\rD_l=\{\pm\chi_i\pm\chi_j\mid1\leq i\neq j\leq l\}\]
and $\rC_l=\rD_l\cup\{\pm2\chi_i\mid1\leq i\leq l\}$. Then we immediately see that $\alpha+\beta\in\rC_l\setminus\rD_l$ exactly means that either $\{\alpha,\beta\}$ or $\{-\alpha,-\beta\}$ equals $\{\chi_i-\chi_j,\chi_i+\chi_j\}$. Then for some $\sigma$ from the Weyl group of $\rD_l$ we get 
\[ \{\sigma\alpha,\sigma\beta\}=\{\chi_{l-1}-\chi_l, \chi_{l-1}+\chi_l\}=\{\alpha_{l-1},\alpha_l\} \]
(the horns of the Dynkin diagram $\rD_l$). Another condition on $\alpha$ and $\beta$ equivalent to $\alpha+\beta\in\rC_l\setminus\rD_l$ is that they do not lie in any subsystem of $\rD_l$ of type $\rA_m$ for $m\geq4$.
\end{rem}

@@

\begin{lemma}
\label{welldef}
Take $d$, $e\leq n$, $d+e=n+1$, $a\in R_d$, $b\in R_e$, and consider an embedding $\iota\colon\rA_4\hookrightarrow\rD_l$. For the chosen simple roots $\alpha_k$ of $\rA_4$ denote $\beta_k=\iota(\alpha_k)$. Then holds
\[ [x_{\beta_1}(a),\,x_{\beta_2+\beta_3}(b)]=[x_{\beta_1+\beta_2}(Nabt\inv),\,x_{\beta_3}(t)], \]
where $N=N_{\beta_1,\beta_2}N_{\beta_2,\beta_3}$.
\end{lemma}
\begin{proof}
Observe that $d$, $e\geq1$. Take $x=x_{\beta_1}(a)$, $y=x_{\beta_2}(N_{\beta_3,\beta_2}bt\inv)$, $z=x_{\beta_3}(t)$. Since $x$ and $z$ lie in the subsystem of type $\rA_4$, we get by R3.1$\!\,_n$ that they commute. Then Hall--Witt identity gives us
\[ \,^y[x,\,[y\inv,\,z]]=\,^x[[x\inv,\,y],\,z], \]
where $[x,\,[y\inv,z]]=[x_{\beta_1}(a),\,x_{\beta_2+\beta_3}(b)]$ and $[[x\inv,y],\,z]=[x_{\beta_1+\beta_2}(Nabt\inv),\,x_{\beta_3}(t)]$ by R2$\!\,_n$. Here $x$ commutes with $x_{\beta_1+\beta_2}(Nabt\inv)$ and $x_{\beta_3}(t)$, and $y$ commutes with $x_{\beta_2+\beta_3}(b)$ by R3.1$\!\,_n$. Finally, $[x_{\beta_1}(-a),\,y]=x_{\beta_1+\beta_2}(Nabt\inv)$ also commutes with $x_{\beta_2+\beta_3}(b)$, what finishes the proof.
\end{proof}

@@

Moreover, with the assumption of Lemma~\ref{welldef} on $a$ and $b$ we have
\[ [x_{\beta}(a),\,x_{\gamma}(b)]=s_{\alpha}(N_{\beta,\gamma}\,ab). \]

The following observation is obvious.
\begin{lemma} \label{commute}
Consider an embedding $\iota\colon\rA_4\hookrightarrow\rD_l$ and denote $\beta_k=\iota(\alpha_k)$ for simple roots $\alpha_k$ of $\rA_4$. Then for any $a\in R_d$, $b\in R_e$, $c\in R_f$ with $d$, $e$, $f\leq n$ holds
\[ [[x_{\beta_1}(a),\,x_{\beta_2}(b)],\,x_{\beta_4}(c)]=1. \]
\end{lemma}

Consider now any two roots $\alpha$, $\beta\in\rD_l$ such that $\alpha+\beta\not\in\rC_l\cup0$. Since $l\geq5$, there exists a subsystem $\Psi$ of type $\rA_4$, containing $\alpha$ and $\beta$. Then we can take any $a\in R_{n+1}$ and $b\in R_d$ for $d\leq n$ and apply Lemma~\ref{commute} for $\Psi$ to get
\begin{equation} \label{sx} [s_\alpha(a),\,x_\beta(b)]=1. \end{equation}
Next, using this fact we can apply again Lemma~\ref{commute} for $\Psi$ and get
\[ [s_\alpha(a),\,s_\beta(c)]=1 \]
for any $c\in R_{n+1}$. Therefore, R3.1$\!\,_{n+1}$ holds for $s_\alpha$'s.

Now one can take $a\in R_d$ for $d\leq0$ and $b\in R_{n+1}$, and repeat the proof of Lemma~\ref{welldef} using~(\ref{sx}) to get
\[ [x_\beta(a),\,s_\gamma(b)]=\begin{cases}s_{\beta+\gamma}(N_{\beta\gamma}\,ab),&d=0,\\x_{\beta+\gamma}(N_{\beta\gamma}\,ab),&d<0.\end{cases} \]
This finishes the proof of R2$\!\,_{n+1}$.

@@

\begin{lemma}
For $a\in R_d$, $b\in R_e$, $d$, $e\leq n$, $d+e=n+1$ holds
\[ [x_{\alpha_{l-1}}(a),\,x_{\alpha_l}(b)]=1. \]
\end{lemma}
\begin{proof}
Take $x=x_{\alpha_{l}-\alpha_{l-2}}(N_{\alpha_{l}-\alpha_{l-2},\alpha_{l-2}}bt\inv)$, $y=x_{\alpha_{l-1}-\alpha_{l-2}}(N_{\alpha_{l-2},\alpha_{l-1}-\alpha_{l-2}}at\inv)$, and $z=x_{\alpha_{l-2}}(t)$. Then
\[ [x_{\alpha_{l-1}}(a),\,x_{\alpha_l}(b)]=[[y\inv,\,z],\,[x,\,z]]=\,^{y\inv z}[y,\,[z\inv,\,x]]\cdot[y\inv,\,[x,\,z]]. \]
Next, by R3.2$\!\,_n$ we know that $x$ and $y$ commute, and therefore 
\[ \,^{y\inv z}[y,\,[z\inv,\,x]]=[[y\inv,\,z],\,x]=x_{\alpha_{l-1}+\alpha_{l}-\alpha_{l-2}}(Nabt\inv) \]
for $N=N_{\alpha_{l-1},\alpha_{l}-\alpha_{l-2}}N_{\alpha_{l}-\alpha_{l-2},\alpha_{l-2}}$. Finally,
\[ [y\inv,\,[x,\,z]]=x_{\alpha_{l-1}-\alpha_{l-2}+\alpha_l}(N'abt\inv) \]
for $N'=-N_{\alpha_{l-2},\alpha_{l-1}-\alpha_{l-2}}N_{{\alpha_{l-1}-\alpha_{l-2}},\alpha_l}$. Since $N_{\alpha,\beta}$ are the structure constants of the Lie algebra, it follows $N=-N'$ from the Jacobi identity.
\end{proof}
